---
title: "Sample Walk-Through for Determining TRS Status"
author: Dan W Joyce
date: "19th April 2018"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true

---

# Introduction
The script `tabulate_cases.R` -- for tabulating each participant and deciding on treatment resistance and responsive cases -- is long and confusing.  This notebook walks through an example case for a TRS participant, hopefully exposing the logic used by executing each chunk and illustrating the results.

```{r message = FALSE}
require(knitr)
source("preprocess_trajectories.R")
```

# Walk-through for a TRS Case
After executing `preprocess_trajectories.R`, we have three tables :
  
  * `TRS.rx.traj` containing the treatment history for all participants in CATIE
  * `TRS.sofas.traj` containing the history of SOFAS/PSP proxy measures for all participants in CATIE
  * `TRS.sx.traj` containing the symptoms (PANSS) history for all participants in CATIE
  * `allIDs` is a vector containing the CATIE ID numbers for all participants in CATIE such that N = `r length( allIDs )`

As an example to walk-through, participant ID = 1438 is detected as a TRS case and this corresponds to index 12 in `allIDs`

We set `i = 12` to select participant 1438 :
```{r}
i <- 12
```

We reproduce the helper functions here so this walk-through will execute:
```{r}
# -- Some helper functions to make the following big loop easier to read ------------------
  CreateNewRow <- function() {
    return( 
      data.frame( 
        ID = thisID,
        time.inTrial = keyvars$es_day[ which( keyvars$src_subject_id == thisID ) ],    # use end of study day as default
        
        onset.Sx  = NA,
        onset.SOF = NA,
        onset.Rx  = NA,
        
        time.onset.TRS = NA,
        
        time.TRS  = NA,
        
        numAdeq = NA,
        durAdeq = NA,
        totalRx = NA,
        
        status.rx = NA,
        status.sx = NA,
        status.sof = NA,
        status.TRS = NA,
        
        TRS.pos = 0,       # resistance in pos domain
        TRS.neg = 0,       # resistance in neg domain
        
        missing.Sx = 0,
        missing.Rx = 0,
        missing.SOF = 0
      )
    )
  }

# -- Function to check if a given participant is missing Rx, Sx and SOF data (to the extent we cannot)
#    robustly process the participant further
CheckZeroRecords <- function( thisRow, thisRx, thisSx, thisSOF ) {
  # -- first, catch case where patient drops out with no real data
  # -- NB : assumes missing.Rx, missing.Sx and missing.SOF are in global environment.
  
  # -- we do this by looking at Rx trajectory : 
  if ( nrow( thisRx ) < 1 )  { ## no treatments recorded at all
    thisRow$status.rx <- NA
    thisRow$totalRx   <- NA
    noFurtherFlag <- 1    ## tell rest of loop to not bother
    missing.Rx <- c( missing.Rx, thisID )
    thisRow$missing.Rx <- 1
  }
  
  # -- second, catch situation where Sx is missing - so we cannot assess response to treatment
  if ( nrow( thisSx ) < 1 ) {
    thisRow$status.sx <- NA
    noFurtherFlag <- 1    ## tell rest of loop to not bother
    missing.Sx <- c( missing.Sx, thisID )
    thisRow$missing.Sx <- 1
  }
  
  # -- third, catch situation where SOF is missing - so we cannot assess social/occ functioning
  if ( nrow( thisSOF ) < 2 ) {
    thisRow$status.sof <- NA
    missing.SOF <- c( missing.SOF, thisID )
    thisRow$missing.SOF <- 1
  }
  
  return( list( thisRow, noFurtherFlag ) )
  
}

```

# Main Loop

## Retrieve Participant's Data
The first part of the large loop determining TRS cases:
```{r eval = FALSE}
# -- Now track each participant ------------------------------------------------------------------------------------------
  for ( i in 1:length( allIDs ) ) {
```

Is this:
```{r}
     
    # -- for this participant ...
    thisID <- allIDs[i]
    
    # -- fetch Sx, SOF and Rx (stored in TRS.sx.traj, TRS.sofas.traj, and TRS.rx.traj) 
    # -- all output from "TRS_CATIE_v3.R"
    thisSx  <- TRS.sx.traj[ which( TRS.sx.traj$ID == thisID ), ]
    thisSOF <- TRS.sofas.traj[ which( TRS.sofas.traj$ID == thisID ), ]
    thisRx  <- TRS.rx.traj[ which( TRS.rx.traj$ID == thisID ), ]
  
    noFurtherFlag <- 0
  
    # -- build a blank data.frame to store this participant
    thisRow <- CreateNewRow()
```

Resulting in three dataframes for this participant.

The PANSS / symptoms history:
```{r}
kable( thisSx )
```

The SOFAS history:
```{r}
kable( thisSOF)
```

And the treatments history
```{r}
kable( thisRx )
```

Finally, an unpopulated "blank row" for the final dataframe is created which looks like:
```{r}
kable( thisRow )
```


## Check for "Zero Data" Conditions

The next chunk checks for participants where there is no usable data -- for example, if there is missing data for SOF, Sx, or Rx -- and `thisRow` is filled accordingly.  This is performed by the helper `CheckZeroRecords` which also returns a flag `= 1` telling the remainder of the loop not to do any further processing.  

```{r}
    # -- check for "zero" records for this participant
    check.zero    <- CheckZeroRecords( thisRow, thisRx, thisSx, thisSOF )
    noFurtherFlag <- check.zero[[2]]
    thisRow       <- check.zero[[1]]
```

Which for this examplar case results in `noFurtherFlag` = `r noFurtherFlag` because there is data to use in assessing TRS status.

## When Participant Does Have Data

We check `noFurtherFlag == 0` indicating data is present and there's further work to be done on this participant ...

```{r eval = FALSE}
    # -- if we have treatment (Rx) data, then process Sx trajectory
    if ( noFurtherFlag == 0 ) {
```

After checking for the (few) cases where visit days don't make sense, we then order `thisSOF`, `thisSX` and `thisRx` on `visday` column to make sure they are in ascending time order:

```{r}

      # -- there is at least one participant who's last.visit.day in thisRx is later than the thisSx.visday
      #    and thisSOF.visday : capture this, and set last.visit.day equal to the last day of thisSx.visday
      if ( thisRx$last.visit.day[ nrow(thisRx) ] > thisSx$visday[nrow(thisSx)] ) {
        thisRx$last.visit.day[ nrow(thisRx) ] <- thisSx$visday[nrow(thisSx)]
      }
      
      # -- order SOF by visday
      thisSOF <- thisSOF[ order( thisSOF$visday ), ]
      # -- remove duplicated SOF rows
      thisSOF <- thisSOF[ !duplicated( thisSOF ), ]
      
      # -- Same for symptoms 
      thisSx  <- thisSx[ order( thisSx$visday ), ]
      thisSx  <- thisSx[ !duplicated( thisSx ), ]
      # -- and treatments (for which there are never any duplicates)
      thisRx  <- thisRx[ order( thisRx$last.visit.day ), ]
```

Next, we (left) join `thisSx`, `thisSOF` and `thisRx` on `visday`, subsetting only on the columns we need:

```{r}
      # -- join a subset of columns in thisSx and thisSOF with thisRx to compile TRS status 
      temp <- left_join( thisSx, 
                         thisSOF[ , names(thisSOF) %in% c("visday","TRS.sofas.traj") ],
                         by = "visday" )  
      
      temp <- left_join( temp, thisRx[ names(thisRx) %in% c("rxDuration","adeqTrial","cumAdeqRx","cumRx","last.visit.day") ],
                         by = c("visday" = "last.visit.day") )
```

Then, we set flags in `temp` for when symptoms (Sx) first "trigger" above the **absolute** threshold for positive and negative domains and another flag for when **both** social/occupational function (SOF) and Sx trigger -- these are the absolute criteria in the TRRIP concensus.

```{r}
     # --------------- time at which participant *first* triggers thresholds for SOF and Sx and Rx
      temp$onsetSxFlag <- ifelse(   ( temp$TRS.pos == 1 |
                                      temp$TRS.neg == 1 ), 1, 0
      )
      
      # -- flag if the participant reached *BOTH* Sx and SOF criteria
      temp$SOF.and.Sx.TRS <- ifelse( temp$onsetSxFlag== 1 & temp$TRS.sofas.traj == 1, 1, 0 )
```

Resulting in :

```{r}
kable( temp )
```

We now select the rows of `temp` which have cumulative numbers of adequate treatment:

```{r}
      # -- filter on cumAdeqRx
      temp2 <- temp[ which( !is.na( temp$cumAdeqRx ) ), ]
```

Resulting in an auxillary dataframe `temp2` which is helpful later:
```{r}
kable( temp2 )
```

We next store the number of adequate treatment trials, the total duration of adequate treatments and the total number of treatments (adequate or otherwise) respectively:

```{r}
      # -- Store number of adequate trials, duration of adequate trials and total treatments
      numAdeq <- max( thisRx$cumAdeqRx )
      durAdeq <- sum( thisRx$rxDuration[ which( thisRx$adeqTrial == 1 ) ] )
      totalRx <- max( thisRx$cumRx )
```

Using `temp2` we establish if absolute SOF and Sx are triggered together -- this was flagged earlier in `SOF.and.Sx.TRS`:

```{r}

      # flag if ever meets SOF and Sx criteria together
      idxEver <- which( temp2$SOF.and.Sx.TRS == 1)
      # -- idxEver : contains rows idxs for when Sx + SOF > threshold
```

We've extracted a vector of row indices, and then test if the length of this vector is non-zero to decide if there's data to work with. 

The next chunk catches cases where the participant never had Sx and SOF absolute thresholds triggering together (and so, can never be TRS), and populates `thisRow` with relevant values for cases where the participant never reaches the absolute thresholds on Sx or SOF:

```{r eval = FALSE}

      # -- if NEVER has Sx and SOF criteria together then set variables to record non-TRS ...
      if( length( idxEver ) == 0 ) {
                
                idxFirstSx  <- which( temp$onsetSxFlag == 1 )
                idxFirstSOF <- which( temp$TRS.sofas.traj == 1 )
                idxFirstRx  <- which( thisRx$adeqTrial == 1 )
                if( length( idxFirstSx ) == 0 ) {
                  onset.Sx <- NA
                } else {
                  onset.Sx  <- temp$visday[ min( idxFirstSx ) ]
                }
                
                if( length( idxFirstSOF ) == 0 ) {
                  onset.SOF <- NA
                } else {
                  onset.SOF <- temp$visday[ min( idxFirstSOF ) ]
                }
        
                # -- as never reaches criteria for Rx >= 2        
                onset.Rx <- NA
                
                time.onset.TRS <- NA
                time.TRS       <- NA
                
                status.rx  <- NA
                status.sx  <- NA
                status.sof <- NA
                status.TRS <- 1    # censored - never met TRS criteria
                
                TRS.pos = 0        # obviously, cannot meet resistance in pos domain
                TRS.neg = 0        # or in neg domain
      }
```

For participant `r thisID`, `length( idxEver )` is `r length( idxEver)` and we skip the above code and move to:

```{r eval = FALSE}
      # -- if this participant DOES meet Sx and SOF criteria together ... establish TRS status
      if ( length( idxEver ) > 0 ) {
        # -- now, ascertain the BASELINE for PANSS change, then when adeqRx >= 2 and
```

Where the main bulk of the work is done for cases where the participant **did** meet absolute SOF and Sx criteria at some point in the trial.

## Computing TRS-related Variables

First we need the time (the `visday`) at which the participant met the absolute Sx and SOF criteria, and also the `visday` at which they meet the >= 2 adequate treatments (Rx):
```{r}
        # -- compute time at which Sx and SOF were "at risk" -- i.e. above threshold, but NOT YET had adequate treatment
        onset.Sx  <- temp$visday[ min( which( temp$onsetSxFlag == 1 ) ) ]
        onset.SOF <- temp$visday[ min( which( temp$TRS.sofas.traj == 1 ) ) ]
        
        # -- and then time at which adequate trials passes threshold >= 2
        onset.rx.idx <- which( temp$cumAdeqRx >= 2 )
        if ( length( onset.rx.idx ) == 0 ) { 
          # -- in this case, participant NEVER had >= 2 adequate trials
          onset.Rx  <- NA
        } else {
          # -- otherwise, store time and which they completed >= 2 adequate trials
          onset.Rx  <- temp$visday[ min( onset.rx.idx ) ]        
        }
        
        # -- compute time at which Sx and SOF were at risk together (but not necessarily the Adequate Rx criteria met)
        time.onset.TRS <- temp$visday[ min( which( temp$SOF.and.Sx.TRS == 1 ) ) ]
        
```

So, for participant `r thisID`, we establish that they:

  * met absolute SOF criteria at `visday` = `r onset.Sx` (`onset.Sx`)
  * met absolute Sx criteria at `visday` = `r onset.SOF` (`onset.SOF`)
  * met **both** Sx and SOF simultaneously on `visday` = `r time.onset.TRS` (`time.onset.TRS`)
  * and the `visday` at which they had completed 2 adequate treatment trials was `r onset.Rx` (`onset.Rx`)

We now need to track the **change** in symptoms, as per the TRRIP concensus criteria.

We extract the baseline PANSS at the timepoint they met both the absolute Sx and SOF criteria (`time.onset.TRS`)

```{r}
        # -- extract the PANSS scores at time.onset.TRS : this will be the baseline for symptom change measurement
        baselinePANSS <- temp[ which( temp$visday == time.onset.TRS ), c("visday", "adjPos", "adjNeg", "adjGen", "adjTot") ]
        
        # -- extract the PANSS scores for followup 
        # -- this will be after 0, 1 or 2 adequate trials and will be the follow-up time for symptom change measurement
```

Resulting in:
```{r}
kable( baselinePANSS )
```

Notice, we use the **adjusted** PANSS scores (making the lowest value 0) so that the Leucht (2009) formula applies (as TRRIP recommends this adjustment for the percentage change criteria).

Next we check : did the participant have **at least** 2 adequate trials?  We can do this by inspecting `onset.Rx` which if NA, means they did not.
Participant `r thisID` clearly does, as `onset.Rx` = `r onset.Rx`.  

```{r}
        # -- never had 2 adequate trials, so we cannot record they were TRS
        if ( is.na( onset.Rx ) ) {
          # -- but we still require some measurement of symptom change
          # -- if they did not have 2 adequate trials, but numAdeq > 0 ==> after one adequate trial, we record PANSS for response
          if ( numAdeq > 0 ) {
            followupPANSS <- temp[ min( which( temp$cumAdeqRx == 1 ) ), c("visday", "adjPos", "adjNeg", "adjGen", "adjTot") ]
          } else 
            # -- NEVER had even one adequate trial, so use end of trial
            # -- find last day with PANSS scores
            {
              tempPANSS <- temp[
                                  complete.cases( temp[ , c("visday", "adjPos", "adjNeg", "adjGen", "adjTot") ] ),
                                  c("visday", "adjPos", "adjNeg", "adjGen", "adjTot")
                                ]
              followupPANSS <- tempPANSS[ nrow( tempPANSS ), ]
            }
        } else
          # -- Had TWO full adequate trials, and therefore
          # -- use onset.Rx day as the follow up AFTER 2 adequate trials
          { 
            followupPANSS <- temp[ which( temp$visday == onset.Rx ), c("visday", "adjPos", "adjNeg", "adjGen", "adjTot") ]
          }
```

Which gives us the PANSS score for the time at which the participant completes 2 adequate trials `visday` = `r onset.Rx`:

```{r}
kable( followupPANSS )
```

We do some duplication checking to ensure both `followupPANSS` and `baselinePANSS` are single rows:

```{r}
        # -- just incase, clear up duplicate panss rows
        followupPANSS <- followupPANSS[ !duplicated( followupPANSS ), ]
        baselinePANSS <- baselinePANSS[ !duplicated( baselinePANSS ), ]
```

Before computing the percentage change in PANSS after 2 adequate trials (by the Leucht 2009 formula):
```{r}
        # -- percentage change in symptoms : Leucht et al, 2009
        deltaPANSS <- 100 * ( baselinePANSS[ c("adjPos", "adjNeg", "adjTot") ] - followupPANSS[ c("adjPos", "adjNeg", "adjTot") ] ) /
                                followupPANSS[ c("adjPos", "adjNeg", "adjTot") ]
```

Resulting in a percentage change vector:

```{r}
kable( deltaPANSS )
```

The next step is to assess if the total PANSS, positive and negative domains, have changed by >= 20%:
```{r}
        # -- Now, if deltaPANSS >= 20% ==> improvement threshold, therefore NOT TRS on that PANSS domain
        TRS.change.PANSS.criteria <- ifelse( deltaPANSS >= 20, 0, 1 )
```

Yielding a vector:
```{r}
kable( TRS.change.PANSS.criteria )
```

Where a `1` indicates the change in symptoms (after 2 adequate treatment trials) was **not greater than** 20%, so they **do** meet TRS on symptom change criteria, and `0` means the change was greater than 20% (and consequently they **do not** meet TRS criteria of symptom change).

Next, record the time at which SOF, Sx and adequate Rx were reached:

```{r}
        # -- Record time at which SOF, absolute Sx and cumAdeqRx >= 2
        idxTimeTRS     <- which( temp2$SOF.and.Sx.TRS == 1 & temp2$cumAdeqRx >= 2 )
        if ( length( idxTimeTRS ) == 0 ) 
          # -- never became TRS
          { 
            time.TRS       <- NA
          } else 
          # -- became TRS
          {
            time.TRS       <- temp2$visday[ min( idxTimeTRS  ) ]
          }
```

Set status flag indicating this participant met the >= 2 **adequate trials** criteria
```{r}
        # -- record IF EVER met Rx >= 2 criteria
        status.rx  <- ifelse( !is.na( onset.Rx ) & onset.Rx  > 0, 1, 0 )
```

Set status flag recording whether this participant meets the full TRS criteria on **symptom change** which is:

  * Positive **or** negative domain scores have not responded by >= 20%
  * And overall PANSS (total) score has not responded by >= 20%

```{r}
        # -- meets change criteria ? 1 => the < 20% change on total *and* one domain 
        status.sx <- ifelse( 
                                ( TRS.change.PANSS.criteria[3] == 1 ) &   # PANSS total change < 20%
                                ( TRS.change.PANSS.criteria[1] == 1 | TRS.change.PANSS.criteria[2] == 1 ),   # either Pos or Neg domains are resistant
                                1, 0 )
```

Finally, check **all** TRRIP criteria have been met (Sx, SOF and Rx), coding `status.TRS` = 2 for full TRS, and 1 otherwise:

```{r}
        # -- absolute criteria threshold STILL met at followup time ONLY if time.TRS is *not* NA
        # -- So, check that time.TRS is not NA (==> simultaneously met absolute Sx and Rx conditions)
        # -- AND that the change in symptoms was actually < 20% -- status.sx.delta
        if( !is.na( time.TRS ) & status.sx == 1 ) {
          # -- the SOF + absolute Sx + Rx criteria were ALL met 
          status.TRS <- 2
        } else {
          # -- participant is treatment response / censored
          status.TRS <- 1
        }
          
        status.sof <- ifelse( !is.na( onset.SOF ) & onset.SOF > 0, 1, 0 )
        TRS.pos = TRS.change.PANSS.criteria[1]       # resistance in pos domain 1,0
        TRS.neg = TRS.change.PANSS.criteria[2]       # resistance in neg domain 1,0
        
```

Some book-keeping:

```{r}
      # -- complete thisRow and store
      thisRow$onset.Sx    <- onset.Sx
      thisRow$onset.SOF   <- onset.SOF
      thisRow$onset.Rx    <- onset.Rx
      
      thisRow$time.onset.TRS <- time.onset.TRS
      thisRow$time.TRS       <- time.TRS
      thisRow$status.TRS     <- status.TRS
      
      thisRow$numAdeq     <- numAdeq
      thisRow$durAdeq     <- durAdeq
      
      thisRow$totalRx     <- totalRx
      thisRow$status.rx   <- status.rx
      thisRow$status.sx   <- status.sx
      thisRow$status.sof  <- status.sof
      thisRow$TRS.pos     <- TRS.pos
      thisRow$TRS.neg     <- TRS.neg
```
 
With the final population `thisRow` vector being:
   
```{r}
kable( thisRow )
```
 
 `thisRow` is then appended to `tab.Surv` for later use.